\textls[-0]{\thetitle}

### 2023-03-04 metatheory-condensed.tex
Reply the comments

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}
in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

>>>>

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}\EDJreply{Yes it is. Look at Section 2.1. When he starts reasoning, he actually deals with category with family (instead of the typing rules), where he consider morphism between contexts, which is substitution. But I agree it can be a bit ambiguous to see (because he literally show two version of MLTT without connecting them---one classical typing rule version, the other directly works on Category with Family (CwF)) but since his proof is mainly on Category with Family, so it counts as explicit substitution. But if you are concerned, we can change to Andras Kovacs "Generalized Universe Hierarchies and First-Class Universe Levels". or Andreas Abel's "Decidability of Conversion for Type Theory in Type Theory". The former one directly work on CwF, the latter one include substitution as part of the raw syntax.}
in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

### 2023-03-04 related-work.tex
Reply the comments

%DTC is as a source of inspiration for may approaches
%\cite{delaware2013,delaware2013b,schwaab2013modular,keuchel2013generic,forsta2020}.
The original DTC encoding requires type-level general recursion that
fails the strict positivity check imposed by proof assistants including Coq and Agda.
\citet{delaware2013} introduce \emph{meta-theory à la carte} (MTC):
it overcomes the problem by using Church encodings for
data types and using Mendler-style folds for evaluation,
though it requires \lsti{Set} impredicativity.
Feature composition is automated through heavy use of type classes.
The framework is implemented as a Coq library.
\citet{schwaab2013modular} adapt DTC to Agda by considering a restricted
class of functors that admit least fixed points.
\citet{keuchel2013generic} use datatype-generic programming techniques
for the underlying representation of type-level fixed points and
avoid \lsti{Set} impredicativity.\YZ{What can we say about Intrinsically-Typed Definitional Interpreters a La Carte?}%\cite{vdr2022intrinsically}

>>>>

%DTC is as a source of inspiration for may approaches
%\cite{delaware2013,delaware2013b,schwaab2013modular,keuchel2013generic,forsta2020}.
The original DTC encoding requires type-level general recursion that
fails the strict positivity check imposed by proof assistants including Coq and Agda.
\citet{delaware2013} introduce \emph{meta-theory à la carte} (MTC):
it overcomes the problem by using Church encodings for
data types and using Mendler-style folds for evaluation,
though it requires \lsti{Set} impredicativity.
Feature composition is automated through heavy use of type classes.
The framework is implemented as a Coq library.
\citet{schwaab2013modular} adapt DTC to Agda by considering a restricted
class of functors that admit least fixed points.
\citet{keuchel2013generic} use datatype-generic programming techniques
for the underlying representation of type-level fixed points and
avoid \lsti{Set} impredicativity.\YZ{What can we say about Intrinsically-Typed Definitional Interpreters a La Carte?}\EDJreply{My understanding is,
 it is really an intrinsic typed version of MTC. It uses different set of tools (like containers (a la altenkirch, can be considered as an alternative of W-type) and stuff) but it is still a design pattern paper.
 But good news is, container is closer to Wtype(inductive type) than the endofunctor encoding (classic MTC approach).}
%\cite{vdr2022intrinsically}
